Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-15 demystify.tex
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

@@

\begin{mybox}
This handbook is available at

@@

\item \emph{You are not a good fit to the program}.  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, NSF leave, or already have too many students currently), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

For example, in the 2024 admission cycle, there is a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Thus, AI/NLP faculty might be 
overwhelmed and cannot consider many, even those with excellent profiles.

@@

\section{Member of an underrepresented or LBGTQ+ groups}\label{sec:urm}
A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} and LBGTQ+ group (, e.g., in your SOP). 
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc).

@@

\paragraph{Research Experiences \autoref{sec:undergrad-research}} You might have more research experiences than international students. For example, you likely have many opportunities to do research as an undergraduates, e.g., through REUs and internships.  Highlight such experiences in your application.

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

@@

\begin{mybox}
This document is available at

@@

\item \emph{You are not a good fit to the program}.  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, personal leave, or already have too many students already), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

For example, 2024 admission cycle has a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Consequently, AI/NLP faculty might be overwhelmed and cannot consider many candidates, even those with excellent profiles.

@@

\section{Member of an underrepresented or LBGTQ+ groups}\label{sec:urm}
A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LBGTQ+ group (, e.g., in your SOP). 
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} understandably students get very excited about their upcoming PhD journal and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  Many profs. have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipments to incoming PhD students (e.g., a laptop).

@@

\paragraph{Research Experiences \autoref{sec:undergrad-research}} You might have many opportunities to do research as an undergraduates, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.
